---
title: "Predicting likelihood of death within 5 days of proning using logistic regression"
author: "David Hannon"
format: 
  pdf:
    documentclass: article
    papersize: a4
    fontfamily: charter
    fontsize: 7pt
    toc: false
    number-sections: true
    geometry: 
    - top=30mm
    - bottom=30mm
    - left=30mm
    - right=30mm
editor: 
  markdown: 
    wrap: 72
---

```{r setup}
#| include: false
set.seed(103L)
library(tidyverse)
library(lubridate)
library(caret)
library(DataExplorer)
library(tidymodels)
library(glmnet)
library(gtsummary)
library(corrplot)
library(mice)
```

## Introduction

Some general preprocessing has been employed thus far to create a raw
dataset that can be further prepared for various machine learning
applications. This chapter will outline my attempts to create a logistic
regression model that gives a likelihood of mortality based on a
patient's response to proning.

This will necessitate

1.  isolating the variables that are likely to be involved in predicting
    the outcome
2.  feature engineering if required
3.  exploring correlations, and decorrelating
4.  train/test split
5.  train the model
6.  test the model and evaluate it

## Variable isolation

```{r load_data}
#| include: false
setwd('/Users/davidhannon/Documents/02. Medicine/Med_Programming/00. Patient DB/ai_df/interim_dfs')
load('pre_post_26_initial_missings_removed.Rda')
pre_post_lr <- pre_post_26
rm(pre_post_26)
```

## Preparing the frame for analysis

The precise time of an observation is not important in this context.
Likewise, the total length of stay is not a variable that can be known
when a patient is still in the ICU, so it too will be removed.

```{r}
pre_post_lr <- pre_post_lr %>% 
  select(-c(time_supine, time_prone, los_days, total_proning_sessions))
```

Arterial blood pressure will be removed as we do not have access to the
amount of vasopressors required to support the BP at this time, vastly
reducing the underlying accuracy and utility of this measure.

```{r}
pre_post_lr <- pre_post_lr %>% 
  select(-c(arterial_pressure_systolic_supine, arterial_pressure_systolic_prone))
```

Finally, we will add replace an index of observations.

```{r}
pre_post_lr <- pre_post_lr %>% 
  mutate(observation = 1:nrow(.))
```

BMI will be binned into categories. The original number will be
retained.

```{r}
#| warning: false

pre_post_lr$bmi_category[pre_post_lr$bmi < 18.5] <- 'underweight'
pre_post_lr$bmi_category[pre_post_lr$bmi >= 18.5 & pre_post_lr$bmi < 25] <- 'normal weight'
pre_post_lr$bmi_category[pre_post_lr$bmi >= 25 & pre_post_lr$bmi < 30] <- 'overweight'
pre_post_lr$bmi_category[pre_post_lr$bmi >= 30 & pre_post_lr$bmi < 35] <- 'obese (class 1)'
pre_post_lr$bmi_category[pre_post_lr$bmi >= 35 & pre_post_lr$bmi < 40] <- 'obese (class 2)'
pre_post_lr$bmi_category[pre_post_lr$bmi > 40] <- 'obese (class 3)'

pre_post_lr <- pre_post_lr %>% 
  mutate(bmi_category = factor(bmi_category)) %>% 
  select(-height_cm)
```

## Basic dataset exploration

### Considering the response variable

It is important to determine an appropriate response variable. My
hypothesis is that the risk of death in the next X days/hours can be
predicted by the patient's 'response to a session of prone positioning'.

Three timeframes are considered:

1.  Death within 3 days / 72 hours of the proning session ending
2.  Death within 5 days / 120 hours of the proning session ending
3.  Death within 7 days / 168 hours of the proning session ending

The next sections will seek to identify variables that seem to show
promise in predicting death within these times.

### Death within 120 hours

```{r}
#| echo: false

pre_post_lr %>% 
  select(-c(proning_session,
            patient_id,
            die_in_72, 
            die_in_168,
            heart_rate_supine, 
            heart_rate_prone, 
            ards_risk_factor, 
            pathogenic_factor, 
            adm_location, 
            end_tidal_co2_marquette_supine,
            peep_supine,
            ventilator, 
            outcome, 
            time_between_abg, 
            session_length, 
            observation,
            et_co2_change_absolute,
            resp_rate_change_absolute,
            minute_volume_change_absolute,
            ph_change_absolute,
            ends_with(c('percent', 'prone')))) %>% 
  tbl_summary(by = die_in_120, missing = 'no', 
              statistic = list(all_continuous() ~ '{mean} ({sd})'), 
              label = list(fi_o2_supine ~ 'FiO2 (%)',
                           sa_o2_systemic_supine ~ 'O2 saturation (%)',
                           ph_abg_supine ~ 'pH',
                           resp_rate_supine ~ 'Resp. Rate (bpm)',
                           pa_o2_supine ~ 'paO2 (kPa)',
                           pa_co2_supine ~ 'paCO2 (kPa)',
                           bicarbonate_abg_a_supine ~ 'Bicarbonate (mmol/L)',
                           lactate_abg_supine ~ 'Lactate (mmol/L)',
                           base_excess_vt_supine ~ 'Base Excess (mmol/L)',
                           potassium_abg_supine ~ 'Potassium (mmol/L)',
                           sodium_abg_supine ~ 'Sodium (mmol/L)',
                           ionised_calcium_abg_supine ~ 'Ionised Calcium (mmol/L)',
                           anion_gap_abg_supine ~ 'Anion Gap (mmol/L)',
                           glucose_abg_supine ~ 'Glucose (mmol/L)',
                           total_haemoglobin_supine ~ 'Hb (g/dL))',
                           white_cell_count_supine ~ 'WCC (x10^9/L)',
                           neutrophils_supine ~ 'Neutrophils (x10^9/L)',
                           lymphocytes_supine ~ 'Lymphocytes (x10^9/L)',
                           c_reactive_protein_supine ~ 'CRP (mg/L)',
                           urea_supine ~ 'Urea (mmol/L)',
                           pcre_supine ~ 'Creatinine (µmol/L)',
                           gfr_supine ~ 'GFR (mL/min/1.73m^2)',
                           haematocrit_supine ~ 'Haematocit',
                           platelet_count_supine ~ 'Platelets (x10^9/L)',
                           albumin_supine ~ 'Albumin (g/L)',
                           predicted_weight ~ 'Predicted Bodyweight (kg)',
                           gender ~ 'Gender',
                           age_years ~ 'Age (years)',
                           weight_kg ~ 'Weight (kg)',
                           ards_type ~ 'ARDS subtype',
                           apache_ii ~ 'APACHE II score',
                           bmi ~ 'BMI',
                           sa_o2_change_absolute ~ 'Change in saO2',
                           #ph_change_absolute, ~ 'Change in pH', (causing an uncertain error)
                           hydrogen_ion_absolute ~ 'Change in H+ ion concentration',
                           pa_o2_change_absolute ~ 'Change in paO2',
                           pa_co2_change_absolute ~ 'Change in paCO2',
                           bicarbonate_change_absolute ~ 'Change in bicarbonate',
                           lactate_abg_change_absolute ~ 'Change in lactate',
                           base_excess_change_absolute ~ 'Change in BE',
                           potassium_abg_change_absolute ~ 'Change in potassium',
                           sodium_abg_change_absolute ~ 'Change in sodium',
                           glucose_change_absolute ~ 'Change in glucose',
                           fi_o2_change_absolute ~ 'Change in FiO2 adminstered',
                           peep_change_absolute ~ 'Change in PEEP adminstered',
                           mean_airway_pressure_change_absolute ~ 'Change in Paw',
                           peak_pressure_change_absolute ~ 'Change in peak airway pressure',
                           pfr_change_absolute ~ 'Change in PF ratio',
                           oxy_factor_change_absolute ~ 'Change in oxygenation factor',
                           vent_ratio_change_absolute ~ 'Change in ventilatory ratio',
                           aa_paco2_change_absolute ~ 'Change in Aa gradient',
                           bmi_category ~ 'Classification of bodyweight'
                           )) %>% 
  add_p() %>% 
  add_significance_stars() %>% 
  modify_caption('Comparison between patients who did and did not die within 120 hours') %>% 
  modify_header(label = '**Variable**') %>% 
  modify_spanning_header(all_stat_cols() ~ 'Death within 120 hours of session')
```

```{r}
#| echo: false
#| warning: false

pre_post_lr %>% 
  select(-c(proning_session,
            patient_id,
            die_in_72, 
            die_in_168,
            heart_rate_supine, 
            heart_rate_prone, 
            ards_risk_factor, 
            pathogenic_factor, 
            adm_location, 
            ventilator, 
            outcome, 
            time_between_abg, 
            session_length, 
            observation,
            et_co2_change_absolute,
            resp_rate_change_absolute,
            minute_volume_change_absolute,
            ph_change_absolute,
            ends_with('percent'))) %>% 
  plot_boxplot(by = 'die_in_120', ggtheme = theme_light())
```

One notable features that can be seen in the above series of box-plots
is the high number of outliers.

## Preparing the dataset for Logistic Regression

It is important to:

1.  Choose good predictors
2.  Process outliers
3.  Deal with missing values

### Choosing predictors

After the application of domain-specific knowledge and cross-referencing
by examining the difference in box-plots between patients who did and
did not die in the days following an episode of prone positioning, the
following candidate variables have been identified:

-   `pa_co2_supine`
-   `bicarbonate_abg_a_supine`
-   `bicarbonate_change_absolute`
-   `bicarbonate_change_percent`
-   `total_haemoglobin_supine`
-   `c_reactive_protein_supine`
-   `hydrogen_ion_abg_nmolL_supine`
-   `hydrogen_ion_percent`
-   `aa_gradient_paco2_supine`
-   `aa_paco2_change_absolute`
-   `aa_paco2_change_percent`
-   `pf_ratio_supine`
-   `pfr_change_absolute`
-   `pfr_change_percent`
-   `oxy_factor_supine`
-   `oxy_factor_change_absolute`
-   `oxy_factor_chane_percent`

This is in addition to the following demographics and outcome variables
of interest:

-   `weight`
-   `bmi`
-   `bmi_category`
-   `age_years`
-   `gender`
-   `ventilator`
-   `apache_ii`
-   `ards_type`
-   `outcome`
-   `die_in_72`
-   `die_in_120`
-   `die_in_168`

```{r}
#| echo: false

pre_post_lr_01 <- pre_post_lr %>% 
  select(observation,
         patient_id,
         age_years,
         gender,
         apache_ii,
         ards_type,
         weight_kg,
         bmi,
         bmi_category,
         outcome,
         die_in_72,
         die_in_120,
         die_in_168,
         pa_co2_supine,
         bicarbonate_abg_a_supine,
         bicarbonate_change_absolute,
         total_haemoglobin_supine,
         c_reactive_protein_supine,
         hydrogen_ion_abg_nmolL_supine,
         hydrogen_ion_absolute,
         aa_gradient_paco2_supine,
         aa_paco2_change_absolute,
         pf_ratio_supine,
         pfr_change_absolute,
         oxy_factor_supine,
         oxy_factor_change_absolute)
```

### Outlier detection

In a small dataset, it is best not to remove entire observations because
of outliers. We will consider the outliers and decide whether to retain
them or remove them (and they can be imputed later).

Initially, each variable was interrogated using a boxplot. The
interquartile range was calculated. Datapoints that fell either 1.5
interquartile ranges below the first quartile, or 1.5 interquartile
ranges above the third quartile, were defined as outliers. The following
outliers are identified in this way:

-   `bmi` = 4 outliers (1.2%)

-   `apache_ii` = 4 outliers (1.2%)

-   `pa_co2_supine` = 9 outliers (2.6%)

-   `bicarbonate_abg_a_supine` = 2 outliers (0.6%)

-   `bicarbonate_change_absolute`= 15 outliers (4.4%)

-   `total_haemoglobin_supine` = 2 outliers (0.6%)

-   `c_reactive_protein_supine` = 4 outliers (1.2%)

-   `hydrogen_ion_abg_nmolL_supine` = 16 outliers (4.7%)

-   `hydrogen_ion_absolute` = 21 outliers (6.2%)

-   `aa_gradient_paco2_supine` = no outliers

-   `aa_paco2_change_absolute` = 2 outliers (0.6%)

-   `pf_ratio_supine` = 3 outliers (0.9%)

-   `pfr_change_absolute` = 10 outliers (3.4%)

-   `oxy_factor_supine` = 4 outliers (1.2%)

-   `oxy_factor_change_absolute` = 10 outliers (2.9%)

### Missing values

| Variable                      | Missing (%)                                                                                       |
|-----------------------|-------------------------------------------------|
| `bmi`                         | `r round(sum(is.na(pre_post_lr_01$bmi)) / nrow(pre_post_lr_01) * 100, 1)`                         |
| `bmi_category`                | `r round(sum(is.na(pre_post_lr_01$bmi_category)) / nrow(pre_post_lr_01) * 100, 1)`                |
| `apache_ii`                   | `r round(sum(is.na(pre_post_lr_01$apache_ii)) / nrow(pre_post_lr_01) * 100, 1)`                   |
| `age_years`                   | 0.0                                                                                               |
| `gender`                      | 0.0                                                                                               |
| `ards_type`                   | 0.0                                                                                               |
| `weight_kg`                   | 0.0                                                                                               |
| `outcome`                     | 0.0                                                                                               |
| `die_in_72`                   | 0.0                                                                                               |
| `die_in_120`                  | 0.0                                                                                               |
| `die_in_168`                  | 0.0                                                                                               |
| `pa_co2_supine`               | `r round(sum(is.na(pre_post_lr_01$pa_co2_supine)) / nrow(pre_post_lr_01) * 100, 1)`               |
| `bicarbonate_abg_a_supine`    | `r round(sum(is.na(pre_post_lr_01$bicarbonate_abg_a_supine)) / nrow(pre_post_lr_01) * 100, 1)`    |
| `bicarbonate_change_absolute` | `r round(sum(is.na(pre_post_lr_01$bicarbonate_change_absolute)) / nrow(pre_post_lr_01) * 100, 1)` |
| `total_haemoglobin_supine`    | `r round(sum(is.na(pre_post_lr_01$total_haemoglobin_supine)) / nrow(pre_post_lr_01) * 100, 1)`    |
| `c_reactive_protein_supine`   | `r round(sum(is.na(pre_post_lr_01$c_reactive_protein_supine)) / nrow(pre_post_lr_01) * 100, 1)`   |
| `hydrogen_ion_nmolL_supine`   | `r round(sum(is.na(pre_post_lr_01$hydrogen_ion_abg_nmolL_supine)) / nrow(pre_post_lr_01) * 100, 1)`    |
| `hydrogen_ion_absolute`       | `r round(sum(is.na(pre_post_lr_01$hydrogen_ion_absolute)) / nrow(pre_post_lr_01) * 100, 1)`        |
| `aa_gradient_paco2_supine`    | `r round(sum(is.na(pre_post_lr_01$aa_gradient_paco2_supine)) / nrow(pre_post_lr_01) * 100, 1)`     |
| `aa_paco2_change_absolute`    | `r round(sum(is.na(pre_post_lr_01$aa_paco2_change_absolute)) / nrow(pre_post_lr_01) * 100, 1)`     |
| `pf_ratio_supine`             | `r round(sum(is.na(pre_post_lr_01$pf_ratio_supine)) / nrow(pre_post_lr_01) * 100, 1)`              |
| `pfr_change_absolute`         | `r round(sum(is.na(pre_post_lr_01$pfr_change_absolute)) / nrow(pre_post_lr_01) * 100, 1)`          |
| `oxy_factor_supine`           | `r round(sum(is.na(pre_post_lr_01$oxy_factor_supine)) / nrow(pre_post_lr_01) * 100, 1)`            |
| `oxy_factor_change_absolute`  | `r round(sum(is.na(pre_post_lr_01$oxy_factor_change_absolute)) / nrow(pre_post_lr_01) * 100, 1)`   |

### Treating NAs and outliers

An important issue in modeling is handling both missing values and outliers. Missing values can be treated in a number of ways, but removing any observations that possesses a missing value is not the best <REF>.

In general, imputing these values is best as long as the total amount of data in a variable being imputed is <5%. We will therefore look for variables that have both a low amount of outliers and a low amount of NAs.








This example will eliminate anything requiring height, oxygenation
index, and impute the missing data.

```{r}
# remove variables with too many (>5% total) outliers and NAs

pre_post_lr_test <- pre_post_lr_01 %>% 
  select(-c(bmi, bmi_category, oxy_factor_supine, oxy_factor_change_absolute))
```

```{r}
# begin investigating of this is 'repeated measure' data
# ==================

# 1. how many patients died within 120 hours of an episode of proning
length(unique(pre_post_lr_test$patient_id[pre_post_lr_test$die_in_120 == T]))

# 2. how many episodes of proning resulted in a death within 120 hours
sum(pre_post_lr_test$die_in_120)
```

```{r}
#| include: false

# impute data

# 1. there is <5% data missing? yes
plot_missing(pre_post_lr_test)
sum(is.na(pre_post_lr_test)) / nrow(pre_post_lr_test)


imp_test <- mice(pre_post_lr_test, method = 'cart', m = 50)
full_data <- complete(imp_test)
plot_missing(full_data)

model_test <- glm(data = full_data, formula = die_in_120 ~ aa_gradient_paco2_supine + aa_paco2_change_absolute)

full_data$pred_class <- predict(object = model_test, newdata = full_data, type = 'response')


```
