---
title: "There is no difference in 28 day mortality between patients who retain improvements in oxygenation following the initial session of prone positioning for moderate to severe ARDS"
author: "David M Hannon"
format:
  pdf:
    documentclass: article
    papersize: a4
    linestretch: 1.15
    mainfont: 'IBM Plex Serif'
    sansfont: 'IBM Plex Serif'
    fontsize: 7pt
    toc: false
    number-sections: false
    geometry: 
    - top=30mm
    - bottom=30mm
    - left=30mm
    - right=30mm
editor: 
  markdown: 
    wrap: 72
bibliography: references.bib
csl: american-medical-association.csl
---

```{r setup}
#| include: false

set.seed(3.14)
library(tidyverse)
load(file = '../data/pre_post_changes.Rda')
```

# Introduction

The mortality of patients suffering from severe ARDS can be improved by
being placed in the prone position. This document outlines a basic
analysis of a dataset of patients from the ICU of University Hospital
Galway who were invasively ventilated and placed in the prone position.

# Methods

Patients were classified as 'responders' or 'non-responders' based on a
comparison of their PF ratio before and after their initial session of
prone positioning. If their PF ratio was higher after returning to the
supine position after their initial session of prone positioning, they
were classified as 'responders'. If the PF ratio was less, they were
classified as 'non-responders'. The 28-day-mortality of the patients was
compared between these two groups.

# Results

```{r}
#| include: false

# add if patient responded
pre_post_changes$pf_responder <- if_else(condition = (pre_post_changes$pf_ratio_supine_post - pre_post_changes$pf_ratio_supine) > 0,
                                         true = 'responder', 
                                         false = 'non-responder')

# look at association between 28 day mortality and proning change retention
changes_sess_01 <- pre_post_changes %>% 
  filter(proning_session == 1)

t <- table(changes_sess_01$mortality_28, changes_sess_01$pf_responder)
rownames(t) <- c('alive in 28 days', 'died within 28 days')
```

A total of 127 sessions of prone positioning were analysed. The two
groups were compared with Pearson's Chi-squared test. The results of
initial grouping is below.

\

```{r}
#| echo: false
#| warning: false

t
```

```{r}
#| include: false
chi <-  stats::chisq.test(changes_sess_01$pf_responder, changes_sess_01$mortality_28, correct = F)
```

\

Statistical analysis showed $\chi^2$ (N = 127) = 0.06, $p$ = 0.94.

# Conclusion
There is no significant difference in 28 day mortality between patients who did and did not retain changes to their PF ratio after their initial session of prone positioning.