---
title: "There is no difference in 28 day mortality between patients who retain improvements in oxygenation following the initial session of prone positioning for moderate to severe ARDS"
author: "David M Hannon"
format:
  pdf:
    documentclass: article
    papersize: a4
    linestretch: 1.15
    mainfont: 'IBM Plex Serif'
    sansfont: 'IBM Plex Serif'
    fontsize: 7pt
    toc: false
    number-sections: false
    geometry: 
    - top=30mm
    - bottom=30mm
    - left=30mm
    - right=30mm
editor: 
  markdown: 
    wrap: 72
bibliography: references.bib
csl: american-medical-association.csl
---

```{r setup}
#| include: false

rm(list = ls())
set.seed(3.14)
library(tidyverse)
load(file = '../data/pre_post_changes.Rda')
```

# Introduction

The mortality of patients suffering from severe ARDS can be improved by
being placed in the prone position. This document outlines a basic
analysis of a dataset of patients from the ICU of University Hospital
Galway who were invasively ventilated and placed in the prone position.

# Methods

The data was refined to include patients with moderate to severe ARDS
(as oper the Berlin criteria). Their initial session of prone
positioning was also isolated.

Patients were classified as 'responders' or 'non-responders' based on
whether they retained beneficial changes to either POF ratio or
Ventilatory Ratio after their initial session of prone positioning. If
their PF ratio was higher after proning than before, they were
classified as a 'responder'. If Ventilatory Ratio was lolwer after
proning they were classified as a 'responder'. Otherwise, they were
classified as a 'non-responder'. The 28-day-mortality of the patients
was compared between these two groups.

```{r}
#| include: false

# refine the dataset - proning sessions over 6 hours, and mod-severe ARDS
pre_post_sess1 <- pre_post_changes %>% 
  filter(proning_session == 1,
         time_between_abg >= 0.25,
         pf_ratio_supine <= 20)
```


# Results

## PF ratio

```{r}
#| include: false

# did patient respond?
pre_post_sess1$pf_responder <- if_else(condition = (pre_post_sess1$pf_ratio_supine_post - pre_post_sess1$pf_ratio_supine) > 0,
                                       true = 'responder',
                                       false = 'non-responder')

# see if any signal here
pf_table <- table(pre_post_sess1$mortality_28, pre_post_sess1$pf_responder)
rownames(pf_table) <- c('lived for 28', 'died within 28')
```

A total of 117 sessions of prone positioning were analysed. The two
groups were compared with Pearson's Chi-squared test. The results of
initial grouping is below.

\

```{r}
#| echo: false
#| warning: false

pf_table
```
\

Comparing the groups using Pearson's Chi Squared showed no significant difference.

```{r}
#| echo: false
#| warning: false

stats::chisq.test(pre_post_sess1$pf_responder, pre_post_sess1$mortality_28, correct = F)
```

\

There is no significant difference between the two groups.

## Ventilatory Ratio

```{r}
#| include: false

pre_post_sess1 <- pre_post_changes %>% 
  filter(proning_session == 1,
         time_between_abg >= 0.25,
         pf_ratio_supine <= 20)

# add if patient responded (was the 'retained' VR LOWER than before proning)
pre_post_sess1$vr_responder <- if_else(condition = (pre_post_sess1$ventilatory_ratio_supine_post - pre_post_sess1$ventilatory_ratio_supine) < 0,
                                       true = 'responder',
                                       false = 'non-responder')

pre_post_sess1$vr_responder <- factor(pre_post_sess1$vr_responder)

pre_post_sess1 <- pre_post_sess1 %>% 
  filter(is.na(pre_post_sess1$vr_responder) == F)

# see if any signal here
vr_table <- table(pre_post_sess1$mortality_28, pre_post_sess1$vr_responder)
rownames(vr_table) <- c('lived for 28', 'died within 28')
```

The amount of cases analysed in this case was less than for PF ratio. This is due to the fact that to calculate Ventilatory ratio requires that we know the patient's height (to calculate Predicted Bodyweight), and this was missing in around 25% of cases. A total of 85 cases were analysed.

Initial results show the following:

\

```{r}
#| echo: false

vr_table
```

\

Comparing the groups using Pearson's Chi Squared showed no significant difference.

```{r}
#| echo: false

stats::chisq.test(pre_post_sess1$vr_responder, pre_post_sess1$mortality_28, correct = F)
```

\

# Conclusion

There is no significant difference in 28 day mortality between patients
who did and did not retain changes to their PF or Ventilatory ratio after their initial
session of prone positioning.
