---
title: "An exploration of oxygenation indices in proned patients"
author: "David M. Hannon"
format:
  pdf:
    documentclass: article
    papersize: a4
    linestretch: 1.15
    mainfont: 'IBM Plex Serif'
    sansfont: 'IBM Plex Serif'
    fontsize: 7pt
    toc: false
    number-sections: false
    geometry: 
    - top=30mm
    - bottom=30mm
    - left=30mm
    - right=30mm
editor: 
  markdown: 
    wrap: 72
bibliography: references.bib
csl: american-medical-association.csl
---

```{r setup}
#| include: false

library(tidyverse)
library(readxl)
library(janitor)
library(gt)
library(gtsummary)
```

# Introduction

The Acute Respiratory Distress Syndrome (ARDS) was initially identified
as a clinical entity in the 1960s[@ashbaugh1967]. ARDS consists of
diffuse alveolar damage due to the activation of alveolar macrophages to
release pro-inflammatory cytokines[@martin1999] that attract neutrophils
to the lungs where they damage the alveolar and capillary epithelium by
release of toxic mediators. This leads to the alveoli being filled with
bloody, proteinaceous fluid. Consequently, the surfactant can no longer
support the alveoli[@ware2001]. The end result is that these damaged
alveoli lead to impaired gas exchange, which is the pathophysiolgic
hallmark of ARDS.

Patients with certain clinical conditions are at higher risk for
developing ARDS. These can broadly be grouped into direct lung injury
risk factors like pneumonia, aspiration, pulmonary contusion,
inhalational injury, near drowning etc. (heretofore referred to as
'ARDSp') and indirect lung injury risk factors such as sepsis,
non-thoracic injuries/hemorrhagic shock, pancreatitis, burns,
drugs/toxins, blood transfusions, cardiopulmonary bypass and reperfusion
injury after lung transplant or embolectomy (which will now be referred
to as 'ARDSexp')[@thompson2017]. Severe lung injury secondary to
Covid-19 has certain unique features, and has been termed
'C-ARDS'[@gattinoni2020].

To date, there are no specific drugs or therapies available to directly
treat/prevent ARDS. Mechanical ventilation that aims to protect injured
lungs and minimize Ventilator Induced Lung Injury (VILI), and management
of refractory hypoxaemia, are the keystones in supportive management of
ARDS[@fan2017]. Part of the above can include placing a patient in the
prone (i.e. 'face down') position. This was first described as a therapy
for ARDS in the 1970s[@piehl1976]. The mechanism by which prone
positioning improves oxygenation is multifactorial. It reduces the
ventral to dorsal transpulmonary pressure difference,
ventilation-perfusion mismatch and lung compression[@cornejo2013;
@jozwiak2013]. Other potential physiological effects of prone
positioning include a decrease in proinflammatory cytokines and
improvement in right ventricle dysfunction by preserving pulmonary
circulation[@guérin2014].

The first prospective randomized control trial (known as the 'PROSEVA'
trial) that showed a mortality benefit from prolonged prone positioning
was conducted in France and published in 2013[@guérin2013]. Prone
positioning for at least 12 to 16 hours per day, while administering low
tidal volumes (4-6ml/kg of ideal body weight), is now strongly
recommended in ventilated patients with severe ARDS[@banavasi2021].

Many questions regarding the utility and efficacy of the prone position
remain. An important issue lies in identifying patients who, although
they fit the criteria to undergo prone positioning, are unlikely to
receive a mortality benefit and in whom other therapies may be
effective[@kawakami2022].

# Population

## Initial selection

```{r load_demo_data}
#| include: false
load(file = '../demographics/demographics.Rda')
```

This section shows details of the population involved in this study. The
data was gathered by interrogating the Electronic Health Record system
used in the Intensive Care Unit (ICU) of University Hospital Galway
(UHG). The system was queried to return all patients on record whp
fulfilled the following criteria:

-   over 18 years of age
-   invasively ventilated
-   placed in the prone position whilst invasively ventilated

## Study population

A total of `r nrow(demo_raw)` records were isolated from the servers.
These records date from between 14/07/2013 and 20/03/2022. They
consisted of patients who had been placed in the prone position
secondary to a pulmonary insult (ARDSp), and extrapulmonary insult
(ARDSexp), and Covid-19 (C-ARDS). Patients in the dataset were proned
between `r min(demo_raw$total_proning_sessions)` and
`r max(demo_raw$total_proning_sessions)` times. Full details of these
demographics can be seen overleaf in Table 01.

\newpage

```{r table_01}
#| echo: false
#| warning: false
#| tbl-cap: Study population demographics
#| tbl-cap-location: bottom

theme_gtsummary_journal(journal = 'nejm')

demo_raw %>% 
  select(-c(patient_id,
            pathogenic_factor,
            ards_risk_factor,
            adm_date,
            dc_date_time,
            ventilator)) %>% 
  tbl_summary(by = ards_type,
              statistic = list(all_continuous() ~ "{mean} ({sd})",
                               all_categorical() ~ "{n} ({p}%)"),
              label = list(gender ~ 'Gender',
                           age_years ~ 'Age (years)',
                           height_cm ~ 'Height (cm)',
                           weight_kg ~ 'Weight (kg)',
                           bmi ~ 'BMI',
                           adm_location ~ 'Admitting location',
                           los_days ~ 'LOS (days)',
                           apache_ii ~ 'Apache II',
                           total_proning_sessions ~ 'Proning sessions',
                           outcome ~ 'Outcome'),
              missing = 'ifany',
              missing_text = 'Not recorded') %>% 
  bold_labels() %>% 
  add_overall()
```

\newpage

## Data gathered around prone-positioning

```{r load_phys_data}
#| include: false

load(file = '../data/pre_post_changes.Rda')
```

For each patient, values relating to ventilation and oxygenation were
isolated around every session of prone positioning that took place
whilst the patient was invasively ventilated. The following values were
recorded:

[Ventilation]{.underline}

-   Fraction of inspired oxygen (FiO~2~)
-   Minute Volume (MV), measured in L/min
-   Positive End-Expiratory Pressure (PEEP), measured in cmH~2~O
-   Peak Inspiratory Pressure (PIP), measured in cmH~2~O
-   Mean Airway Pressure (P~aw~), measured in cmH~2~O

[Oxygenation]{.underline}

-   peripheral oxygen saturation (SpO~2~), recorded using bedside pulse
    oximetry
-   saturation of oxygen (SaO~2~), recorded from arterial blood gas
-   partial pressure of oxygen (PaO~2~), recorded from arterial blood
    gas

[Other]{.underline}

-   partial pressure of carbon dioxide (PaCO~2~), recorded from arterial
    blood gas
-   haemoglobin concentration (g/dL), recorded from arterial blood gas

# Oxygenation indices

## Overview

The following indices of oxygenation will be examined. Some have been
recorded directly from monitoring equipment to which the patient was
attached in ICU, and some have been calculated. The indexes are outlined
below. Summary statistics and an overview of the distribution for each
variable follows.

The following recorded *directly*:

-   SpO~2~
-   SaO~2~
-   PaO~2~

The following were *calculated* from recorded values:

-   P/F ratio
-   Ventilatory ratio
-   Oxygenation index
-   Oxygenation factor
-   A-a O~2~ gradient
-   Arterial O~2~ content (CaO~2~)

## P/F ratio

PaO~2~/FiO~2~ ratio is the ratio of arterial oxygen partial pressure to
fractional inspired oxygen. The equation is:

$$ {P/F \ ratio} = \frac {PaO_2}{FiO_2} $$

```{r pf_ratio}
#| include: false
```

## Ventilatory ratio

$$  Ventilatory \ ratio = \frac{MV \times Pa_{CO_2}}{{PBW \times 100 \times 5}}   $$

## Oxygenation index

$$ Oxygenation \ index = \frac{FiO_2 \times P_{AW}}{PaO_2}  $$

## Oxygenation factor

$$ Oxygenation \ factor = \frac{PF \ ratio}{P_{AW}}  $$


## A-a O~2~ gradient

$$ A-a \ gradient = [FiO_2 \times (P_{atm} - P_{H_2O}) - \frac{Pa_{CO_2}}{0.8}] - Pa_{O_2}$$

## Arterial O~2~ content

$$ Ca_{O_2} = (1.34 \times [Hb] \times Sp_{O_2}) \ + \ (0.0225 \times Pa_{O_2})  $$


# Explorations & Comparisons

\newpage

# References
